---
title: "CAA2018 Tutorial First Part"
author: "Martin Hinz"
date: "2 März 2018"
output: 
  html_document:
    keep_md: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Basic Calibration in R
### Available Options

* [ArchaeoPhases](https://cran.rstudio.com/web/packages/ArchaeoPhases/index.html)
  * [Marie-Anne Vibet](#https://avatars3.githubusercontent.com/u/15785508?s=40&v=4) Marie-Anne Vibet
  * "Post-Processing of the Markov Chain Simulated by 'ChronoModel', 'Oxcal' or 'BCal'"
  * Not really for calibration itself, has to be done externally
* [ArchSeries](https://github.com/davidcorton/archSeries)
  * [David Orton](https://avatars3.githubusercontent.com/u/7224976?s=40&v=4) David Orton
  * "Frequency time-series tools for archaeology"
  * No calibration options
  * Not on CRAN yet
* [Bchron](http://cran.rstudio.com/web/packages/Bchron/index.html)
  * [Andrew Parnell](https://avatars3.githubusercontent.com/u/3535973?s=40&v=4) Andrew Parnell
  * "A Bayesian radiocarbon chronology model with R"
  * BchronCalibrate()
* [rcarbon](https://cran.r-project.org/web/packages/rcarbon/index.html)
  * [Andrew Bevan](https://avatars3.githubusercontent.com/u/3984941?s=40&v=4) Andrew Bevan
  * "Calibration and Analysis of Radiocarbon Dates"
  * calibrate()
* [oxcAAR](https://cran.r-project.org/web/packages/oxcAAR/index.html)
  * [ISAAK](https://avatars3.githubusercontent.com/u/19709572?s=40&v=4) ISAAK
  * "Interface to 'OxCal' Radiocarbon Calibration"
  * oxcalCalibrate()
* Hand made (it is actually not that hard)
  * Basic
  * by means of matrix algebra (see [my blog post on basic calibration](http://martinhinz.info/jekyll/update/blog/2016/06/03/simple_calibration.html)) 
  * by means of bayesian statistics (see [my blog post on bayesian calibration](http://martinhinz.info/jekyll/update/blog/2017/01/23/bayesian_calibration.html))

Essentially it is Bchron, rcarbon and oxcAAR!

## Installing the packages with calibration option

So lets start with them and install them if necessary

```{r install_packages}
# we might need the devtools package to install other packages from Github
library(devtools)

# if not already done, install c14bazAAR and oxcAAR from Github

isaak_packages <- c("c14bazAAR", "oxcAAR")

devtools::install_github(
  repo = c("ISAAKiel/c14bazAAR", "ISAAKiel/oxcAAR")
)

library(oxcAAR) # load package
library(c14bazAAR) # load package

# A bit of magick
# I added dplyer because we might need it later
package_list <- c("Bchron", "rcarbon", "dplyr") # The packages we would like to have

for (package in package_list){ # for each in the list
  if( !is.element(package, .packages(all.available = TRUE)) ) { #check if it is not already installed
    install.packages(package, repos = "http://cran.us.r-project.org") # if not, install it
  }
  library(package,character.only = TRUE) # load package
}
```

Luckily, all calibration functions are named differently, so there is no overwriting of the functions going on here. Only dplyr is masking some functions, but this should not bother us. We can have all of them side by side.

## Just doing calibration

We might be eager to do our first calibration, so lets start immediatly. We need a 14C date for that, consisting of BP uncal and BP standard deviation:

```{r introducing_14_c_date}
bp <- 4000
std <- 50
```

With that at hand, we can start calibration:

### rcarbon

Here it is straight forward, just plug in your date and let it run

```{r quick_cal_rcarbon}
calDate.rcarbon <- calibrate(bp,std)
str(calDate.rcarbon)
```

The result is a list containing
* $metadata: The metadata for the (uncalibrated) date
* $grids: The actual calibrated probabilities
* $calmatrix: A matrix of probability values, one row per calendar year (BP!) in timeRange and one column per date. Just a different format of $grids. This defaults to NA if not `calMatrix=TRUE` is specified in the call.

So let's plot it 
```{r quick_cal_rcarbon_plot}
plot(calDate.rcarbon)
```

Nice and clean!

### Bchron

Again straight forward, specify BP, std and (here necessary, in rcarbon optional) the calibration curve

```{r quick_cal_BChron}
calDate.Bchron <- BchronCalibrate(ages = bp, ageSds = std, calCurves = "intcal13")
str(calDate.Bchron)
```

The result is a list containing
* $ages: The uncalibrated date
* $ageSds: The standard deviation of the uncalibrated date
* $calCurves: The calibration curve used
[We have specified this as input parameters]
* $ageGrid: The calibrated BP (!) values
* $densities: The calibrated probabilities

So let's plot it 

```{r quick_cal_BChron_plot}
plot(calDate.Bchron)
```

Also nice, lacking the uncal probability and the calibration curve, but highlighting the 95% highest density region (2 sigma range). Note that time is running here from right to left (BP dates).

### oxcAAR

oxcAAR needs a bit of extra love before it works. It relies on OxCal as calibration backend, so we have to install it first. Luckily there is a helper for doing that:

```{r oxcAAR_install_OxCal}

quickSetupOxcal()
oxcAAR::setOxcalExecutablePath("~/OxCal/bin/OxCalLinux")

```

Oxcal is installed in your working directory, and the path to the executable is set. With that we might proceed to the calibration:

```{r quick_cal_oxcAAR}
calDate.oxcAAR <- oxcalCalibrate(bp, std)
str(calDate.oxcAAR)
```

The resulting object is a bit more complicated. It is a list of calibrated dates (one each for each uncalibrated date inserted in the call), each containing

* $name: A name for the date. If not given, a consecutive number.
* $bp, $std: The uncalibrated date and its standard deviation
[We have specified this as input parameters]
* $cal_curve: The calibration curve used, with full parameterisation. Defaults to intcal13
* $sigma_ranges: The sigma ranges for the calibrated date (1,2,3 sigma)
* $raw_probabilities: a data frame containing the BC (!) dates and the calibrated probabilities

Again, let's plot it 

```{r quick_cal_oxcAAR_plot}
plot(calDate.oxcAAR)
```

Sigma ranges, probabilities, everything is in there.

Want a calibration curve?

```{r quick_cal_oxcAAR_calcurve_plot}
calcurve_plot(calDate.oxcAAR)
```

### Multiple dates

So, if you like to calibrate multiple dates, that is also not hard:

```{r cal_multiple_dates}
calMultiDate.rcarbon <- calibrate(x = c(4000,3000), errors = c(50,25) )
str(calMultiDate.rcarbon)
calMultiDate.Bchron <- BchronCalibrate(ages = c(4000,3000), ageSds = c(50,25), calCurves = c("intcal13","intcal13"))
str(calMultiDate.Bchron)
calMultiDate.oxcAAR <- oxcalCalibrate(bp = c(4000,3000), std = c(50,25))
str(calMultiDate.oxcAAR)

```

In case of Bchron, for every date a calibration curve must be given independently. Can be a bit tedious, but one can script that, too.

The result is all the same like when calibrating an individual date, but in every case two results are produce, one for each uncalibrated date. In case of rcarbon, the dates are internally separated, in case of Bchron and oxcAAR, the result is simply is list of calibrated dates. 

How are multiple dates visualised in each package?

```{r cal_multiple_dates_plot}
plot(calMultiDate.rcarbon)
plot(calMultiDate.Bchron)
plot(calMultiDate.oxcAAR)
```

In case of Bchron there is an individual plot for each date, with oxcAAR both dates are plotted on the same x-axis (time), while rcarbon only returns the first date. With an additional parameter you can convince rcarbon to plot the second date, but all the time only one date per plot.

```{r cal_multiple_dates_plot_rcarbon_plot_second_date}
plot(calMultiDate.rcarbon, ind = 2)
```

## Why oxcAAR?

### Differences between calibration algorithms

In archaeology, OxCal has become a quasi standard for calibration. One might like that or not, but to make results comparable it is actually not bad to agree on a standard. Calibration itself is not a hard thing to do (we will see that later), but the results of the different implementations might vary. Lets compare (in doing so we shift the dates from oxcAAR by 1950 to make them BP and add 0.5 due to the specific output values of OxCal):

```{r comp_results_packages, echo=F}

plot(calDate.rcarbon$grids$`1`$calBP,calDate.rcarbon$grids$`1`$PrDens, type="l",ylab = "probability", xlab = "cal BP")
lines(calDate.Bchron$Date1$ageGrid,calDate.Bchron$Date1$densities, col="red")
lines(1950 - calDate.oxcAAR$`1`$raw_probabilities$dates + 0.5,
      calDate.oxcAAR$`1`$raw_probabilities$probabilities, col="green")
legend("bottomright", c("rcarbon", "Bchron", "oxcAAR"), col = c("black", "red", "green"), title = "calibration method", lty=1)
```

While rcarbon and oxcAAR are quite similar (based on the same algorithm, although OxCal works on 5 years resolution), Bchron results differ.

If we zoom in and just look at the differences:

```{r comp_results_packages_detail, echo=F}
# 1. Join the results on a common time axis
## 1.1 Gather rcarbon and Bchron results

all_dates <- c(calDate.Bchron$Date1$ageGrid, calDate.rcarbon$grids$`1`$calBP)
my_start <- max(all_dates)
my_end <- min(all_dates)

my_date_sequence <- my_start:my_end

my_compare_df <- data.frame(date = my_date_sequence)
my_compare_df$Bchron <- full_join(my_compare_df, as.data.frame(calDate.Bchron$Date1), by = c("date" = "ageGrid"))$densities

my_compare_df$rcarbon <- full_join(my_compare_df, calDate.rcarbon$grids$`1`, by = c("date" = "calBP"))$PrDens

## 1.2. oxcAAR result has to be abit shifted due to specific OxCal output
calDate.oxcAAR_for_compare <- calDate.oxcAAR$`1`$raw_probabilities
calDate.oxcAAR_for_compare$dates <- 1950 - calDate.oxcAAR_for_compare$dates + 0.5

my_compare_df$oxcAAR <- left_join(my_compare_df, calDate.oxcAAR_for_compare, by = c("date" = "dates"))$probabilities

# oxcal has 5 years resolution, so only use those dates
my_compare_df <- my_compare_df[my_compare_df$date%%5==0,]
my_compare_df$rcarbon <- my_compare_df$rcarbon - my_compare_df$oxcAAR
my_compare_df$Bchron <- my_compare_df$Bchron - my_compare_df$oxcAAR
my_compare_df$oxcAAR <- 0

plot(my_compare_df$date,my_compare_df$oxcAAR, col="green", ylim = c(-6e-04,6e-04), type="l",ylab = "Difference from OxCal Result", xlab = "cal BP")
lines(my_compare_df$date,my_compare_df$rcarbon, col="black")
lines(my_compare_df$date,my_compare_df$Bchron, col="red")
legend("bottomright", c("rcarbon", "Bchron", "oxcAAR"), col = c("black", "red", "green"), title = "calibration method", lty=1)
```

Also rcarbon differs slightly from the implementation in OxCal. This might not mean much, and it does not imply that OxCal is actually the correct result, but it means that the different algorithm produce slighly different probabilities.

So, if you like to work with results that are exactly like those who are using OxCal, oxcAAR might be the best option. This especially holds true if you later want to do rather esoteric things like sum calibration. Removing every possible source of mistrust might be relevant in this field, still.

### Sequential calibration

What is also possible with oxcAAR (and as it seems, with no other package currently), is using the capabilities of OxCal to produce sequential calibrations. This is an aspect that should be made more accessible in the future, but it is already possible now.

Lets assume we have something like that:

```{oxcal eval=FALSE}
top soil
------------------------
  Phase 3
Sample 6 BP 3100 +- 25
Sample 5 BP 3250 +- 25
------------------------
  Phase 2
Sample 4 BP 3230 +- 25
Sample 3 BP 3370 +- 25
------------------------
  Phase 1
Sample 2 BP 3640 +- 25
Sample 1 BP 3340 +- 25
------------------------
  virgin soil
```

We could represent this in a dataframe like this:
```{r prepare_sequence_calibration}
my_sequence <- data.frame(name=character(), phase=numeric(), bp=numeric(), std = numeric(), stringsAsFactors = F)

my_sequence[1,] <- list("Sample 6", 3, 3100, 25)
my_sequence[2,] <- list("Sample 5", 3, 3250, 27)
my_sequence[3,] <- list("Sample 4", 2, 3230, 31)
my_sequence[4,] <- list("Sample 3", 2, 3370, 50)
my_sequence[5,] <- list("Sample 2", 1, 3640, 28)
my_sequence[6,] <- list("Sample 1", 1, 3340, 33)

my_sequence
```

In the real world you might have the data already present in a convenient format, like csv data exported from Excel.

```{r make_string_for_sequence_calibration}

# make a list of strings, an element per phase with the R_Dates
phased_dates <- by(my_sequence,my_sequence$phase, function(x)
  R_Date(x$name,x$bp,x$std))

# Wrap the dates in Phases
phases <- Phase(names=c("1","2","3"), r_dates_strings = phased_dates)

# make an vector with some names for the boundaries
boundary_name <- c("begin","1->2","2->3","end")

# wrap the phases in boundaries
my_sequence_elements <- wrap_in_boundaries(phases, boundary_name)

# Wrap the whole thing in a sequence
my_oxcal_code <- Sequence(my_sequence_elements, name = "my_sequence")

cat(my_oxcal_code)
cat()
# call Oxcal and read and parse the result
my_result_file <- executeOxcalScript(my_oxcal_code)
my_result_text <- readOxcalOutput(my_result_file)
my_result_data <- parseOxcalOutput(my_result_text, only.R_Date = F)

my_result_data
plot(my_result_data)
```

As you see, now we have the dates calibrated according to the specified model. The unmodelled distributions are given in lighter, the modelled in darker gray. Also the boundaries are displayed.

The graphical output might improve soon, but already now all the essential informations from OxCal can be plotted, and probably more important, accessed via R.
