---
title: "CAA2018 Tutorial Second Part"
author: "Clemens Schmid"
date: "2 März 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## c14bazAAR

c14bazAAR is a R package to download and prepare bulk c14 date collections.

### Why is this useful?

#### Radiocarbon dates

- c14 dates are fairly standardised (a pillar of stability in the Bad Data of archaeology)
- c14 dates have high resolution temporal and spatial information (they can be linked to all kind of data -- archaeology, climatology, genetics, linguistics...) 
- c14 dates are linked to distinct events in human history (they tell stories)

#### An R package?

- User perspective
    - access many highly different databases with one interface
    - reproducibility with scripted data selection
    - standard data structures for direct access to powerful R tools (tidyverse)
- Developer perspective
    - Open Source: examine & improve the implementation and adjust everything for your needs
    - simple parser development framework to add further databases
    - embed bulk c14 dates into your own application

### Open Archives

- **14SEA** 14C database for Southeast Europe and Anatolia (10,000–3000 calBC).
- **aDRAC** Archives des datations radiocarbone d'Afrique centrale by Dirk Seidensticker.
- **AustArc** A Database of 14C and Luminescence Ages from Archaeological Sites in Australia by Alan N. Williams, Sean Ulm, Mike Smith, Jill Reid
- **CALPAL** Radiocarbon Database of the CalPal software package by Bernhard Weninger. See nevrome/CalPal-Database for an interface.
- **CONTEXT** Collection of radiocarbon dates from sites in the Near East and neighboring regions (20.000 - 5.000 calBC) by Utz Böhner and Daniel Schyle.
- **EUROEVOL** Cultural Evolution of Neolithic Europe Dataset by Katie Manning, Sue Colledge, Enrico Crema, Stephen Shennan and Adrian Timpson.
- **CARD Upload Template - KITE East Africa v2.1** Radiocarbon dates from eastern Africa in the CARD2.0 format by Colin Courtney Mustaphi, Rob Marchant
- **RADON** Central European and Scandinavian database of 14C dates for the Neolithic and Early Bronze Age by Dirk Raetzel-Fabian, Martin Furholt, Martin Hinz, Johannes Müller, Christoph Rinne, Karl-Göran Sjögren und Hans-Peter Wotzka.
- **RADON-B** Database for European 14C dates for the Bronze and Early Iron Age by Jutta Kneisel, Martin Hinz, Christoph Rinne.
- **...**

If you know more, add them in [this](https://github.com/ISAAKiel/c14bazAAR/issues/2) issue or join the development of c14bazAAR and start an own pull request.


```{r cars}
summary(cars)
```

