---
title: "CAA2018 Tutorial Third Part"
author: "Martin Hinz"
date: "2 März 2018"
output:
  html_document:
    keep_md: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Mass Operations on 14C Data

Having received the dates from the c14bazAAR, we might like to analyse the demographic development of Baden-Würtemberg between 7500 and 2700 cal BP (~ 5500–700 cal BC). For doing so, one first step might be to produce a sum calibration of those dates.

```{r}
library(dplyr)
library(oxcAAR)
library(magrittr)
quickSetupOxcal()

# Taking the output of the c14bazAAR and transform them into OxCal code.

bawue_final <- readRDS(file = "bawue_final.RDS")


bawue_dates_for_oxcaar <- R_Date(bawue_final$labnr,
                                 bawue_final$c14age,
                                 bawue_final$c14std) 

bawue_sum_code <- oxcal_Sum(bawue_dates_for_oxcaar) # Wrap it into a sum command
```

Looking to the beginning of that code:

```{r}
cat(substr(bawue_sum_code,1,253))
```

Looks like proper OxCal code. Than, lets use it for sum calibration

```{r}
my_result_data <- bawue_sum_code %>%
  executeOxcalScript() %>%
  readOxcalOutput() %>%
  parseOxcalOutput(only.R_Date = FALSE, first = TRUE)

str(my_result_data)
```

So, we get back a object of oxcAARCalibratedDatesList with one entry, the sum. Lets plot that:

```{r}
plot(my_result_data)
```

Nice wiggly and wobbly. But how much can we trust that pattern? Could it come just from the shape of the calibration curve, or from random effects from the date distribution? Lets find out using simulation!

## Using simulations to qualify sum calibration

For constructing confidence intervalls using Monte Carlo methods (simulations), we at first have to agree on the null hypothesis. The so called 'UCL approach' is to make a generalised linear regression of the data and use this (a kind of smooth general trend) as null model.

We make this somewhat easier, but nevertheless also very relevant given the appropriate research question: Does the distribution of the data deviate significantly from an uniform random distribution?

For that we need a uniform random distribution of our dates, and from that we simulate ^14^C dates.

In oxcAAR, there is a simple to use function to simulate a 14C sum. It is called `oxcalSumSim()`. 

```{r}
simsum <- oxcalSumSim(
  n = nrow(bawue_final),
  stds = sample(bawue_final$c14std),
  timeframe_begin = -5500,
  timeframe_end = -700,
  date_distribution = "uniform"
)

plot(simsum)
```

Now we have kind of a background, which we can compare to our real world data. In an ideal world, the plot would be just a flat plateau, since we introduced uniform distributed original data. But due to random distribution effects and those of the calibration curve, there are still wiggles, indicating that there is some kind of variability going on. We might just compare to our real sum calibration:

```{r}
# for the sake of simplicity, we extract the raw probabilities of our result
my_sum_result <- my_result_data$` Sum `$raw_probabilities

plot(
  my_sum_result$dates,
  my_sum_result$probabilities,
  type="l"
) # The reals world in solid line
lines(
  simsum$raw_probabilities$dates,
  simsum$raw_probabilities$probabilities,
  lty=2
  ) # The simulation in dashed line
```

There are some wiggles of the real word data that are much higher than the simulated ones, some are much lower. There seems to be a pattern that differs from our hypothesis of a uniform distributed real world. But how significant is that. Have in mind, if I repeat the simulation, we get another, different result:

```{r}
simsum2 <- oxcalSumSim(
  n = nrow(bawue_final),
  stds = sample(bawue_final$c14std),
  timeframe_begin = -5500,
  timeframe_end = -700,
  date_distribution = "uniform"
)

plot(
  simsum$raw_probabilities$dates,
  simsum$raw_probabilities$probabilities,
  type="l",
  lty=2
  ) # The simulation in dashed line
lines(
  simsum2$raw_probabilities$dates,
  simsum2$raw_probabilities$probabilities,
  lty=3
  ) # The second simulation in dotted line
```

These differences only come from the different distribution of the dates, and at second order, on the different position on the calibration curve. We can estimate the intensity of this effects, if we repeat the simulation several times. Lets say, for the sake of run time, we do that 10 times:

```{r}
library(broom)
library(ggplot2)

sim_result <- bawue_final %>% bootstrap(10) %>% do(
  oxcalSumSim(
    n = nrow(bawue_final),
    stds = sample(bawue_final$c14std),
    timeframe_begin = -5500,
    timeframe_end = -700,
    date_distribution = "uniform"
  )$raw_probabilities
)

ggplot() + geom_line(data=sim_result, aes(x=dates,y=probabilities,group=replicate))
```

Lets zoom in abit

```{r}
time_window <- c(-5500, -700)

ggplot() + geom_line(data=sim_result, aes(x=dates,y=probabilities,group=replicate)) + xlim(time_window)
```

Now we have some variety, and we can add our real world data to it to see whether or not its intensity is exceeding the simulation results:

```{r}
last_plot() + geom_line(data = my_sum_result, aes(x=dates,y=probabilities), color="red")
```

Some ranges of the real world data definitively exceeds the random scatter of the simulation. But lets make that more standardised, and compute the 95% interval of the simulation. If the real world data goes beyond or below this range, than there is only a 5% chance that this could have occured by chance, making it statistically significant (of course for this you would rather go to 1000 repetitions of the simulation):

```{r}
alpha = .05

envelope <- sim_result %>% group_by(dates) %>% summarize(low=quantile(probabilities, alpha / 2),
                                         high=quantile(probabilities, 1 - alpha / 2))

ggplot() +
  geom_ribbon(data = envelope, aes(x=dates,ymin=low, ymax = high), alpha = .5) + 
  geom_line(data = my_sum_result, aes(x=dates,y=probabilities), color="red") + xlim(time_window)
  
```

Now we get an idea. But there are still all this tiny wiggles, and most people (eg. Williams 2012) suggest using smoothing to get rid of them. We also do that! OxCal results come usually with a 5 years resolution, so if we apply a moving window average with a with of 10, we get a 50 years smoothing window.

```{r}
ma <- function(x,n=10){stats::filter(x,rep(1/n,n), sides=2)} # The smoothing function

my_sum_result %<>% mutate(probabilities_smoothed = ma(probabilities))

ggplot() +
  geom_ribbon(data = envelope, aes(x=dates,ymin=low, ymax = high), alpha = .5) + 
  geom_line(data = my_sum_result, aes(x=dates,y=probabilities), color="red") + 
  geom_line(data = my_sum_result, aes(x=dates,y=probabilities_smoothed), color="blue") + xlim(time_window)
  
```

And also for the envelope:

```{r}
envelope %<>% mutate(low_smoothed = ma(low), high_smoothed = ma(high))

ggplot() +
  geom_ribbon(data = envelope, aes(x=dates,ymin=low_smoothed, ymax = high_smoothed), alpha = .5) + 
  geom_line(data = my_sum_result, aes(x=dates,y=probabilities_smoothed), color="blue") + xlim(time_window)
```

Those that are brave might now repeat this with a little more simulation runs, lets say 1000 (takes approx. 1h, dependend on your computer):

```{r cache=T, eval=F}
sim_result <- bawue_final %>% bootstrap(1000) %>% do(
  oxcalSumSim(
    n = nrow(bawue_final),
    stds = sample(bawue_final$c14std),
    timeframe_begin = -5500,
    timeframe_end = -700,
    date_distribution = "uniform"
  )$raw_probabilities
)

envelope <- sim_result %>% group_by(dates) %>% summarize(low=quantile(probabilities, alpha / 2),
                                         high=quantile(probabilities, 1 - alpha / 2))

my_sum_result %<>% mutate(probabilities_smoothed = ma(probabilities))

envelope %<>% mutate(low_smoothed = ma(low), high_smoothed = ma(high))

ggplot() +
  geom_ribbon(data = envelope, aes(x=dates,ymin=low_smoothed, ymax = high_smoothed), alpha = .5) + 
  geom_line(data = my_sum_result, aes(x=dates,y=probabilities_smoothed), color="blue") + xlim(time_window)
```

The result shows where the pattern in the real world curve can not be explained by the effects of random sample distribution or effects of the calibration curve. This does not mean that the result is correct: **Garbage in, Garbage out**! You still have to make sure (as sure as possible) that your sample is representative!
